\documentclass[12pt letter]{report}
\input{./template/preamble}
\input{./template/macros}
\input{./template/letterfonts}

\title{\Huge{Network Layer: Data Plane}}
\author{\huge{Madiba Hudson-Quansah}}
\date{}
\usepackage{parskip}

\setcounter{tocdepth}{4}
\setcounter{secnumdepth}{4}

\begin{document}
\maketitle
\newpage
\pdfbookmark[section]{\contentsname}{too}
\tableofcontents
\pagebreak

\chapter{Introduction}

The role of the network layer is to move segments, encapsulated into
datagrams from the source to the destination. The network layer is
responsible for forwarding datagrams from one router to the next
until the destination is reached. The network layer also provides
logical addressing, which is used to identify devices on the network
and to route datagrams to their destination.

\section{Forwarding and Routing: Data and Control Planes}

The network layers has two primary functions:
\begin{description}
  \item[Forwarding]  - The process of moving packets form the input
    input link of a router to the appropriate output link. Forwarding
    is the local action taken by a router when it receives a packet.
    The router examines the packet's header, determines the
    appropriate output link based on its forwarding table, and then
    forwards the packet to the next hop. This is implemented in the
    Data Plane of the network layer.
  \item[Routing] - The process of determining the route or path taken
    by packets as they flow from sender to receiver. The algorithms
    that calculate these paths are called \textbf{routing algorithms}
    an they determine the path along which packets flow from source
    router to destination router. This is implemented in the Control
    Plane of the network layer.
\end{description}

% \section{Control Plane}
%
% \subsection{Traditional}
%
% \subsection{SDN Approach}

\section{Network Service Model}

The \textbf{Network Service Model} defines the characteristics of
end-to-end delivery of packets between sending and receiving hosts.

The network layer defines different types of services that it can
offer, these include:
\begin{description}
  \item[Guaranteed Delivery] - Guarantees that a packet sent by a
    source host will eventually arrive at the destination host.
  \item[Guaranteed Delivery with Bounded Delay] - Guarantees delivery
    within a specified host-to-host delay bound
  \item[In-Order Packet Delivery] - Guarantees that packets arrive at
    the destination in the order they were sent by the source.
  \item[Guaranteed Minimal Bandwidth] - Guarantees that a certain
    minimum bandwidth will be
    available for the transmission of packets between the source and
    destination.
  \item[Security] - Provides mechanisms to ensure the
    confidentiality, integrity, and authenticity of data transmitted
    between the source and destination.
\end{description}

The Internet's  network layer provides a single \textbf{best-effort
service}, i.e.:
\begin{itemize}
  \item Packets are not guaranteed to be delivered
  \item Packets not guaranteed to be delivered in the order they were sent
  \item There is no guaranteed on the end-to-end delay of packet delivery
  \item There is no guarantee on the minimum bandwidth available for packet
    transmission
\end{itemize}

\chapter{Router Internals}

The components of a router are as follows:
\begin{description}
  \item[Input Ports] - Responsible for receiving packets from the
    incoming links, performing the following functions:
    \begin{itemize}
      \item Physical layer function of terminating an incoming
        physical link at the router.
      \item Link Layer functions needed to interoperate with the link
        layer at the other side of the incoming link, such as
        framing, error detection, and possibly link-layer flow control.
      \item The lookup function that determines the output port to
        which the packet should be forwarded, based on the
        destination IP address in the packet's header and the
        forwarding table. It is here where the appropriate output
        port is determined and the packet is forwarded to switching fabric.
    \end{itemize}
  \item[Switching Fabric] - Connects the router's input ports to its
    output ports.
  \item[Output Ports] - Stores packets received from the switching
    fabric and transmits these packets on the outgoing link by
    performing the necessary link-layer, and physical layer functions.
  \item[Routing Processor] - Performs control-plane function, such as
    executing routing protocols, maintaining routing tables and
    attached link state information, and computing forwarding tables,
    traditionally. In Software Defined Networking (SDN), the routing
    processor is responsible for communicating with the SDN
    controller to receive forwarding rules and updates.
\end{description}

Data plane functions are usually implemented in hardware for
high-speed packet processing, while Control plane functions are
typically implemented in software, allowing for flexibility and ease of updates.

Depending on the type of forwarding used different information is
required to be stored in the forwarding table, for example:
\begin{description}
  \item[Destination-Based Forwarding] - The forwarding table contains
    entries that map destination IP addresses to output ports. When a
    packet arrives, the router looks up the destination IP address in
    the forwarding table to determine the appropriate output port for
    forwarding the packet.
  \item[Generalized Forwarding] - The forwarding table contains entries that map
    various packet header fields (such as source and destination IP
    addresses, protocol type, port numbers, etc.) to specific
    actions (such as forwarding to a particular output port, dropping
    the packet, or modifying the packet header). This allows for more
    flexible and granular control over packet forwarding based on
    multiple criteria.
\end{description}

\section{Input Port Processing and Destination-Based Forwarding}

Again the input port is where the forwarding table is consulted to
determine the appropriate output port for forwarding the packet, via
the switching fabric. The forwarding table is either computing and
updated by the routing processor, using a routing protocol, or is
received from a remote SDN controller.

\dfn{Prefix}{
  A prefix is a contiguous sequence of bits in an IP address that
  represents a network or subnet. In the context of IP routing, a
  prefix is used to identify a group of IP addresses that share a common network
  portion. For example, in the IP address range
  \begin{align*}
    11001000\quad 00010111\quad 00010000 \quad 00000000 \\
    \text{through} \\
    11001000\quad 00010111 \quad 00010111 \quad 11111111
  \end{align*}
  The prefix would be the first 21 bits, i.e.:
  \begin{align*}
    11001000\quad 00010111\quad 00010
  \end{align*}
}

In Destination-Based forwarding the router matches a \textbf{prefix}
of the packet's destination address with the entries in the
forwarding table, if there's a match, the router forwards the packet
to a link associated with the match.

Because its possible for multiple entries in the forwarding table to match the
destination address, the router uses the \textbf{longest prefix
match}, i.e. the entry with the longest matching prefix is chosen for
forwarding the packet.

This process of matching the destination address with the forwarding table
and sending it to the appropriate output port is a specific case of
the general \textbf{Match, and Action} abstraction model performed in
many networked devices.

\section{Switching}

\dfn{Non-blocking}{
  A packet being forwarded to an output port will not be blocked from
  reaching that output port as long as no other packet is currently
  being forwarded to that output port.
}

The switching fabric is responsible for actually switching/forwarding
packets from an input port to an output port. This can be done in a
number of ways, i.e.:
\begin{description}
  \item[Switching via Memory] - Switching between input and output
    ports is done under the direct control of the CPU (routing
    processor), with input and output ports functioning as
    traditional I/O devices in an operating system. An input port
    with an arriving packet signals the routing processor via an
    interrupt, the packet is then copied into processor memory, where
    the routing processor extracts the destination address, and
    looked up the appropriate output port in the forwarding table and
    is then copied to the output port's buffers.
    In this scenario if the memory  bandwidth is such that a maximum
    of $B$ packets per second can be written into or read from, then
    the overall forwarding throughput from input ports to output
    ports must be less than $\displaystyle{\frac{B}{2}}$. This also
    comes with the caveat that two packets cannot be forwarded at the
    same time even of they have different destination ports as only one memory
    access can be done at a time over the shared system bus.
  \item[Switching via a Bus] - An input port transfer a packet
    directly to the output port over a shared bus, without the
    involvement of the routing processor. This is sufficient for a
    small local area and enterprise networks where there is a small
    number of ports and low traffic volume. However, as the number of
    ports and traffic volume increases, the shared bus can become a
    bottleneck, limiting the overall throughput of the router.
  \item[Switching via Interconnection Networks] - An interconnection
    network is a high-speed switching fabric that connects the input
    ports to the output ports. A crossbar switch is an
    interconnection network consisting of $2N$ buses that connect $N$
    input ports to $N$ output ports, which each bus connecting one
    input port to one output port. This allows for multiple packets
    to be switched simultaneously, as long as they are destined for
    different output ports. The crossbar switch can achieve a maximum
    throughput of $N$ packets per second, as each of the $N$ input
    ports can be connected to a different output port at the same time.
    A crossbar switch is \textbf{non-blocking}, however of two
    packets are destined for the same output port, one of the packets
    will be blocked until the output port is free. For larger routers
    with more ports and higher traffic volumes, more complex
    interconnection networks, such as multi-stage interconnection
    networks, may be used to provide higher throughput and scalability.
\end{description}

\section{Output Port Processing}

Output port processing takes packets that have been stored in the
output port's memory and transmits them over the output link, this
includes scheduling, and dequeuing packets for transmission, and
performing the necessary link-layer and physical layer functions to
transmit the packet over the outgoing link.

\section{Queuing}

Packet queuing can occur at both the input and output ports of
router, with the location and extent of which depending on the
traffic load, the relative speed of the switching fabric and the line
speed of the outgoing link.

\textbf{Packet Loss} occurs when there is no more memory available to
store arriving packets, this can occur at both the input and output
ports of a router, and is typically due to congestion in the network.
When a router's input or output port is congested, it may drop
incoming packets, leading to packet loss.

\subsection{Input Queuing}

If the switching fabric is not fast enough, relative to input line
speeds, to transfer all arriving packets through the fabric without
delay, then packets will queue up at the input ports, waiting or
their turn to be transferred through the switch fabric.

Head-of-Line (HOL) blocking occurs when the first packet in the input
queue is destined for an output port that is currently busy, while subsequent
packets in the queue are destined for different output ports that are
free. In this scenario, the first packet blocks the subsequent
packets from being forwarded, even though they could be forwarded to
their respective output ports. This can lead to increased latency and
reduced throughput in the router, as packets are unnecessarily
delayed due to HOL blocking.

\subsection{Output Queuing}

If the switching fabric is faster than the input line speeds, but the
output line speeds are slower than the switching fabric, then packets
will queue up at the output ports, waiting for their turn to be
transmitted over the outgoing link.

\dfn{Drop-Tail}{
  A simple packet dropping policy where incoming packets are dropped
  when the queue is full, i.e. when there is no more memory available
  to store incoming packets.
}

\dfn{Drop-Front}{
  A packet dropping policy where the packet at the front of the queue
  is dropped to make room for an incoming packet when the queue is full.
}

When there is not enough memory to buffer an incoming packet, a
decision must be made to either drop the arriving packet, i.e.
\textbf{drop-tail}, or to remove one already in the queue to make
room for the arriving packet, i.e. \textbf{drop-front}.

Active Queue Management (AQM) algorithms can be used to proactively
manage the queue and prevent congestion, an example is the Random
Early Detection (RED) algorithm, which randomly drops incoming
packets with a certain probability when the average queue size
exceeds a certain threshold, in order to signal congestion to the
sender and encourage it to reduce its sending rate.

A consequence of output queuing is that a \textbf{packet scheduler}
at the output port must choose one packet among those queued for transmission.

\subsection{Buffering}

The right buffer size is crucial for the performance of a router, as
it can help to mitigate the effects of congestion and prevent packet
loss. Given that
the buffer size is $\mbold{B}$, the average round-trip time is,
$\mbold{RTT}$, the link capacity is $\mbold{C}$, and the number of
concurrent TCP connections is $\mbold{N}$, then the recommended
amount of buffering is given by the formula:
\[
  \mbold{B} = \mbold{RTT} \times \frac{C}{\sqrt{\mbold{N}} }
\]

Large buffers are not necessarily always better as they could lead to
longer queuing delays, which can negatively impact the performance of
real-time applications and increase the overall latency in the
network. Additionally, large buffers can lead to a phenomenon known
as \textbf{bufferbloat}, where excessive buffering causes high
latency and jitter in the network, leading to poor performance for
applications that require low latency, such as online gaming and
video conferencing.

\section{Packet Scheduling}

\subsection{First-in-First-Out (FIFO)}

In this approach packets arriving at the output port are transmitted
in the order the arrived in. If there is no buffer space to hold the
arriving packet the queue's packet-discarding policy is applied, i.e.
either drop-tail or drop-front.

\subsection{Priority Queuing}

Packets arriving at the output port are given priorities based on
certain criteria such as packet type i.e., network management
information packets vs user traffic, real-time vs non real-time, etc.
Each priority level has its own queue and the scheduler will transmit
from the highest priority queue in a FIFO manner, then the next
highest level and so on.

\subsection{Round Robin and Weighted Fair Queuing (WFQ)}

Packets are sorted into classes based on certain criteria, but rather
than having a FIFO ordering, it alternates among the classes,
transmitting one packet from each class in a round-robin manner.

A \textbf{work conserving queue} will never allow the link to remain
idle when there are packets of any class are waiting to be
transmitted, i.e. such a scheduler will look for a packet in a given
class, if the class is empty it immediately moves on to the next
class until it finds a non-empty class, and then transmits a packet
from that class.

A generalized for of round robin queuing is the \textbf{Weighted Fair
Queuing} approach, where arriving packets are classified and queued
in the appropriate per-class waiting area, then serviced in a
circular manner, i.e. class 1 to class 2 to class 3, etc. coming back
to class 1 and so on. WFQ differs from round robin in that each class
may receive a differential amount of service in any interval of time.

Specifically, each class $i$, is assigned a weight $w_i$, during any
interval of time during which there are class $i$ packets to send,
class $i$ will then be guaranteed to receive a fraction of service equal to
\[
  \displaystyle\frac{w_i}{\displaystyle\Sigma w_j}
\]
Where the sum in the denominator is taken over all classes that also
have packets queues for transmission. In the worst case, even if all
classes have queued packets class $i$ will still be guaranteed to
receive the fraction $\displaystyle\frac{w_i}{\Sigma w_j}$ of the
bandwidth, where in the worst case the sum in the denominator is over
all classes.

\chapter{Internet Protocol (IP)}

\section{IPv4 Datagram Format}

The IPv4 protocol is comprised of the following fields:
\begin{description}
  \item[Version number]  - 4 bits, specifies the IP protocol version
    of the datagram. Allows the router to determine how to interpret
    the remainder of the IP datagram.
  \item[Header Length] - 4 bits, the length of the entire header as
    it can vary. Helps in determining the start and end of the data payload.
  \item[Type of Service] - 8 bits, allows different types of IP
    datagrams to be distinguished from each other, .e.g. real-time
    datagrams from non real-time ones.
  \item[Datagram Length] - 16 bits, The total length of the IP
    datagram (header + data) in bytes.
  \item[Identifier, flags, fragmentation offset] - 16, 3, 13 bis
    respectively, relates to IP fragmentation, when a large IP
    datagram is broken up into several smaller IP datagrams which are
    then forwarded independently to the destination where they are reassembled.
  \item[Time-to-live] - 8 bits, ensures that datagrams do not
    circulate forever in the network
  \item[Protocol] - 8 bits, used only when an IP datagram reaches its
    final destination to indicate the specific transport layer
    protocol to which the data portion of the datagram should be
    passed, e.g. TCP, UDP, etc.
  \item[Header checksum] - 16 bits, used for error detection of the IP header.
  \item[Source and Destination IP addresses] - 32 bits each, used to
    identify the source and
    destination hosts of the datagram.
  \item[Options] - 32 bits, optional field that can be used for
    various purposes, such as security, routing, and network testing.
    The options field is not commonly used in practice and is often left empty.
  \item[Data] - Variable length, contains the payload of the datagram, which
    can be up to 65,535 bytes in size (including the header).
\end{description}

\section{IPv4 Addressing}

\dfn{Interface}{
  The boundary between a host and the physical link to which it is
  attached, or the boundary between a router and the physical link to
  which it is attached.
}

A router has multiple interface, one for each of its links. Because
every host and router is capable of sending and receiving  IP
datagrams, IP requires each host and router interface to have its own
IP address, thus an IP address is technically associated with an
interface rather than a host or router as a whole.

Each IP address is 32 bits (4 bytes) long, and there are a total of
$2^{32}$ possible unique public IP addresses. IP addresses are
displayed in dotted-decimal notation where each byte of the address
is written in its decimal form and is separated b a period from other
bytes, i.e:
\begin{center}
  10.10.0.1
\end{center}
In binary form the above IP address would be represented as:
\begin{align*}
  00001010 \quad 00001010 \quad 00000000 \quad 00000001
\end{align*}

\subsection{Subnetting}

\dfn{Subnet}{
  A subnet is a logical subdivision of an IP network, created by
  dividing a larger network into smaller, more manageable segments. Each
  subnet is identified by a unique subnet mask, which is used to
  determine the network and host portions of an IP address.

}

\dfn{Subnet Mask}{
  A subnet mask is a 32-bit number that is used to divide an IP address
  into a network and host portion. It is typically represented in
  dotted-decimal notation, similar to an IP address. The subnet mask
  consists of a series of consecutive 1s followed by a series of
  consecutive 0s, where the number of 1s indicates the length of the
  network portion of the IP address, and the number of 0s indicates
  the length of the host portion.
}

A portion of an interface's IP address is determined by the
\textbf{subnet} to which it is connected. Hosts connected to the same
subnet have the same prefix in their IP addresses, and can
communicate directly with each other without the need for a router.
The subnet mask indicates that the left most $n$ bits of the IP
address define the subnet address. For example, if the subnet mask is
223.1.1.0/24, then the leftmost 24 bits of the IP address define the
subnet address, and the remaining 8 bits define the host address
within that subnet.

Subnets can also exist between routers, in this case the interfaces
of the routers that are connected to the same subnet will have the
same subnet address, and can communicate directly with each other
without the need for additional routing.

To determine the subnets of a network, detach each interface from its
host/router, creating islands of isolated networks, with interfaces
terminating the end points of the isolated networks.

The Internet's address assignment strategy, is known as Classes
Interdomain Routing (CIDR), which generalizes the notion of subnet
addressing. The 32 bit IP address is divided into two parts and has
the dotted decimal notation of the form:
\begin{align*}
  a.b.c.d/x
\end{align*}
Where $x$ indicates the number of bits in the prefix. The $x$ most
significant bits of an address in this form constitute the network
portion of the IP address, with an organization typically assigned a
block of contiguous addresses, i.e., a range of addresses with a common
prefix.

The remaining 32-$x$ bits of n address can be thought as
distinguishing among the devices within the organization all of which
have the same network prefix.

Before CIDR, the network portions of an IP address were constrained
to be 8, 16, or 24 bits long, known as \textbf{classful addressing},
since subnets with 8-, 16- and 24-bit subnet addresses were known as
class A, B and C respectively. CIDR allows for more flexible
subnetting, as the network portion of an IP address can be of any
length, allowing for more efficient use of the available IP address
space and better support for hierarchical routing.

The broadcast address 255.255.255.255 is used when a message is to be
delivered to all hosts on the same subnet, optionally forwarding the
message into other subnets, but not to hosts on other subnets.

\subsection{Dynamic Host Configuration Protocol (DHCP)}

The Dynamic Host Configuration Protocol (DHCP) is a network protocol
that allows hosts to obtain/be allocated an IP address automatically,
via a DHCP server. A network administrator can configure a DHCP
server  so that a host is given the same IP address each time it
connects to the network, or a \textbf{temporary IP address}, that
changes each time the host connects to the network.

DHCP is often referred to as a \textbf{plug-and-play} / \textbf{zero
configuration (zeroconf)} protocol. The process by which a host is
allocated an IP address via DHCP is as follows:
\begin{description}
  \item[DHCP Servre Discovery]  - A newly arriving host sends a DHCP
    discover message via UDP on the network, since the host does not
    have an IP address yet it uses the broadcast address to send the
    message, and the source IP address is set to 0.0.0.0.
  \item[DHCP Server Offer] - DHCP servers on receiving a DHCP
    discover message responds with a DHCP offer message that is broadcast
    to all nodes again. Since several DHCP servers can be present on
    the subnet the client may find itself receiving multiple DHCP
    offer messages, each containing a different IP address offer from
    the different DHCP servers. Each server off message contains the
    transaction ID of the received discover message, the proposed IP
    address of the client, the network mask ,and an IP address lease
    time, which is the amount of time for which the client can use
    the proposed IP address before it must renew the lease.
  \item[DHCP request] - The client selects one of the DHCP offer
    messages and responds with a DHCP request message echoing back
    the configuration parameters, again broadcasted to all.
  \item[DHCP ACK] - The server responds to the DHCP request message
    with a DHCP ACK message, confirming the requested parameters.
\end{description}

\section{Network Address Translation (NAT)}

With the rapid growth of the Internet and the limited availability of
IPv4 addresses, the availability of public IP addresses has become a
significant concern. Network Address Translation (NAT) is a technique
used to address this issue by allowing multiple devices on a private
network to share a single public IP address when accessing the Internet.

A NAT-enabled router has an interface that is part of the home
network, and addressing within the home network is done using private
IP addresses, which are not routable on the public Internet. When a
device on the home network wants to access the Internet, the NAT
router translates the private IP address of the device into the
public IP address of the router, allowing the device to communicate
with external servers on the Internet.

A NAT-enabled router maintains a translation table used to determine
the actual host inside the private network to which an incoming
packet should be forwarded, this table is updated each time a new
outgoing connection is made from the private network to the Internet,
and is used to translate the public IP address back to the private IP
address when a response is received from the Internet.

\section{IPv6}

IPv6 is a 128 bit successor to the IPv4 protocol, designed to address
the limitations of IPv4, such as:
\begin{itemize}
  \item Limited address space as IPv4 only support $2^{32}$ addresses
    while IPv6 supports $2^{128}$ addresses, providing a much larger
    address space to accommodate the growing number of devices
    connected to the Internet.
  \item Improved security features, such as built-in support for IPsec, which
    provides confidentiality, integrity, and authentication for IP
    communications.
  \item  Better support for Quality of Service (QoS) by allowing for
    more efficient routing and traffic management, which is important
    for real-time
    applications such as video streaming and online gaming.
\end{itemize}

\subsection{IPv6 Datagram Format}

The IPv6 datagram format is similar to that of IPv4, but with some changes:

\begin{description}
  \item[Expanded addressing capabilities]  - IPv6 uses 128-bit
    addresses, allowing for a much larger address space compared to
    IPv4's 32-bit addresses.
  \item[40-byte header] - The IPv6 header is fixed at 40 bytes in
    length, compared to the variable-length header of IPv4, which can
    be up to 60 bytes long.
  \item[Flow labeling] - IPv6 includes a flow label field in the
    header, which can be used to identify and prioritize specific
    flows of traffic, allowing for better support for real-time applications.
\end{description}

\subsection{Transitioning from IPv4 to IPv6}

Tunnelling is a technique used to facilitate the transition from IPv4
to IPv6 by encapsulating IPv6 packets within IPv4 packets, allowing
them to be transmitted over an IPv4 network infrastructure. This
allows for the gradual deployment of IPv6 while still maintaining
compatibility with existing IPv4 networks. The intervening set of
IPv4 routers between two IPv6 routers are referred to as the
\textbf{tunnel}, and the IPv4 addresses of the two IPv6 routers are
referred to as the \textbf{tunnel endpoints}.

\chapter{Generalized Forwarding and Software Defined Networking (SDN)}

In generalized forwarding a match-plus-action table generalizes the
forwarding table, as forwarding desertions are made based on the
match of multiple header fields, such as source and destination IP
addresses, protocol type, port numbers, etc., rather than just the
destination IP address as in destination based forwarding. This
allows for more flexible and granular control over packet forwarding
based on multiple criteria.

The OpenFlow standard is a widely used protocol for implementing
software-defined networking (SDN), and relies on a match-plus-action
table known as a \textbf{flow table}, this table includes:
\begin{description}
  \item[A set of header field values] -
\end{description}

\section{Match}

\section{Action}

\section{OpenFlow}

\chapter{Middleboxes}

\end{document}
