\documentclass[12pt letter]{report}
\input{./template/preamble}
\input{./template/macros}
\input{./template/letterfonts}

\title{\Huge{Introduction}}
\author{\huge{Madiba Hudson-Quansah}}
\date{}
\usepackage{parskip}

\setcounter{tocdepth}{4}
\setcounter{secnumdepth}{4}

\begin{document}
\maketitle
\newpage
\pdfbookmark[section]{\contentsname}{too}
\tableofcontents
\pagebreak

\chapter{Fundamentals}

\section{Eight Great Ideas}

\begin{itemize}
  \item Design for Moore's Law
  \item Use abstraction to simplify design
  \item Make the common case fast
  \item Performance via parallelism
  \item Performance via pipelining
  \item Performance via prediction
  \item  Hierarchy of memories
  \item Dependability via redundancy
\end{itemize}

\subsection{Design for Moore's Law}


\subsection{Abstraction}

\subsection{Semiconductors}

\dfn{Semiconductor}{
  A material that has an electrical conductivity between that of a conductor and an insulator.
}

\dfn{P-Type}{
  A semiconductor material that has been doped with an electron acceptor element. This creates a material that has a net positive charge.
}

\dfn{N-Type}{
  A semiconductor material that has been doped with an electron donor element. This creates a material that has a net negative charge.
}

\dfn{PNP Transistor}{
  A type of transistor that consists of a layer of N-type semiconductor between two layers of P-type semiconductor.
}

\dfn{NPN Transistor}{
  A type of transistor that consists of a layer of P-type semiconductor between two layers of N-type semiconductor.
}

Semiconductors are used in the construction of transistors, which are the building blocks of modern computers.
Transistors are made up of three parts:
\begin{itemize}
  \item Gate
  \item Drain
  \item Source
\end{itemize}


\end{document}
